---
title: "Bargraphs for One and Two Grouping Variables"
author: "Tabitha Njoroge"
date: "3/29/2020"
output: html_document
---
### 1. Set the chunk options
```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 4,
	fig.width = 12,
	message = FALSE,
	warning = FALSE
)
```

### 2. Load the required packages
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
```

### 3. Read in the required datasets
```{r}
df = read.csv("C:/Users/Peter/Desktop/Wambui-Rstudio/rjourney/datasets/wafanyikazi.csv")
data = read.csv("C:/Users/Peter/Desktop/Wambui-Rstudio/rjourney/datasets/training_data.csv")


```
### 4. Generating a bargraph
#### 4.1 Bargraph for one grouping variable
<br>
4.1.1 Department Distribution Bargraph
<b>
```{r}
##Generate the required dataset
dept  = df %>% 
  group_by(Department) %>% 
  summarise(count = n())

knitr::kable(dept)

### Generate the bargraph               
dept_bargraph = ggplot(data = dept, aes (x = Department, y = count)) +
                geom_bar(stat = "identity",fill = "red", width = 0.7) +
                geom_text(aes(label = count), hjust = 0.5, vjust = -0.25) +
                theme(plot.title = element_text(face = "bold",
                                                hjust = 0.5, 
                                                size = 16),
                      plot.subtitle = element_text(face = "bold",size = 16,
                                                   hjust = 0.5),
                      plot.caption = element_text(face = "italic",size = 10,
                                                  colour = "blue"),
                      axis.title = element_text(face = "bold", size = 14),
                      axis.text = element_text(face = 12, size = 12),
                      axis.line = element_line(size = 1.3),
                      panel.background = element_rect(fill = "NA")) +
                  labs(title = "Department Distribution",
                       subtitle = "Barplot",                        
                       caption = "Figure 4.1.1",
                       x = "Department",
                       y = "Frequency")
dept_bargraph
```

*Give the theme of the bargraph an object name for purposes of reptitive use

```{r}
bargraph_theme = theme(plot.title = element_text(face = "bold", 
                                                 hjust = 0.5,size = 16),
                 plot.subtitle = element_text(face = "bold",size = 8, 
                                           hjust = 0.5),
                 plot.caption = element_text(face = "italic",size = 10, 
                                          colour = "blue"),
                 axis.title = element_text(face = "bold", size = 14),
                 axis.text = element_text(face = 12, size = 12),
                 axis.line = element_line(size = 1.3),
                 panel.background = element_rect(fill = "NA"))
```

<b>
<b>
4.1.2 Average Income per County
<b>
```{r}
##Generate the required dataset
average_county_income = df %>% 
  group_by(County) %>%
  summarise(average_income = round(mean(Income),0))

knitr::kable(average_county_income)

### Generate the bargraph
county_bargraph = ggplot(data = average_county_income, aes(x = County, y = average_income))+
                  geom_bar(stat = "identity", fill = "cyan", width = 0.7)+
                  geom_text(aes(label = average_income), hjust = 0.5,
                            vjust = -0.25, size = 3)+
                  bargraph_theme+
                  labs (title = "Average Income per County", 
                        caption = "Figure 4.1.2",
                        x = "County",
                        y  = "Average Income")
county_bargraph

```

#### 4.2 Bargraph for two grouping variable
<br>
4.2.1 Average age per department and per role
<b>
```{r}
##Generate the required dataset
age_dept_role = df %>% 
  group_by(Department, Role) %>% 
  summarise(average_age = round(mean(Age),0))

knitr::kable(age_dept_role)

#Generate the graph
dept_role_bg = ggplot(data = age_dept_role, aes(x = Department, 
                                                y = average_age, 
                                                fill = Role))+
               geom_bar(stat = "identity", position = "dodge") +
               geom_text(aes(label = average_age), hjust = 0.5, 
                         vjust = -0.25, 
                         position = position_dodge(width = 0.9)) +
               bargraph_theme+
               theme(legend.title = element_blank())+
               scale_fill_manual(values = c("purple","pink","cyan"))+
               labs(title = "Average Age \n per \n Department per Role", 
                    caption = "Figure 4.2.1",
                          x = "Role",
                          y = "Average Age")
dept_role_bg

```

#### 5. FSDT Data by Zindi
<b>
Gender distribution in the three age groups
```{r}
###Gender Count per Age group
##Generate the required dataset
tab1 = data %>% 
  select(ID, Q1, Q2) %>% 
  mutate(gender = ifelse(Q2 == 1, "Male", "Female")) %>% 
  rename(age = Q1) %>% 
  mutate(age_groups = ifelse(age >= 15 & age <=35, "Youth",
                       ifelse(age >= 36 & age <= 65, "Mid-age",
                              "Senior Citizens"))) %>% 
  group_by(age_groups, gender) %>% 
  summarise(age_groups_count = n())

knitr::kable(tab1)

##Generate the bargraph
gender_age_bargraph = ggplot(data = tab1, aes(x = age_groups, 
                                              y = age_groups_count, 
                                              fill = gender))+
                      geom_bar(stat = "identity", position = "dodge")+
                      geom_text(aes(label = age_groups_count), hjust = 0.5,
                                vjust = -0.25, 
                                size = 3, color = "red", 
                                position = position_dodge(width = 0.7))+
                      theme(legend.title = element_blank())+
                      bargraph_theme +
                      scale_fill_manual(values = c("purple", "pink"))+
                      labs(title = "Gender Distribution per Age group", 
                           caption = "FSDT 2017; Fig 5.1",
                           x = "Age_groups", 
                           y = "Count")
gender_age_bargraph 
```
<b>
Gender distribution on 'Money from Savings Interest / Investments'.
<b>
```{r}
## Generate the required data
gender_income = data %>%
  select(ID, Q2, Q8_6) %>%
  mutate(gender = ifelse(Q2 == 1, "Male", "Female")) %>%
  mutate(invest_income = ifelse(Q8_6 == 1, "Yes", "No")) %>%
  group_by(gender, invest_income) %>%
  summarise(Count = n()) %>%
  mutate(perc = round((Count/sum(Count))*100,1))

knitr::kable(gender_income)


## Generate the bar graph
gender_income_bargraph = ggplot(data = gender_income, aes(x = invest_income,
                                                          y = perc, 
                                                          fill =gender))+
                         geom_bar(stat = "identity", position = "dodge",
                                  width = 0.6)+
                         geom_text(aes(label = paste0(perc, "%")), 
                                   hjust = 0.5, vjust = -0.3,
                                   size = 3,
                                   position = position_dodge(width = 0.7))+
                         bargraph_theme+
                         theme(legend.title = element_blank())+
                         scale_y_continuous(labels = function(x) 
                           paste0(x, "%"))+
                         scale_fill_manual(values = c("purple","pink"))+
            labs(title = "Gender Distribution in Savings/Investment Income",
                 subtitle = "(Gender percentage on income from savings interests,investments,stocks,unit trusts etc.)",
                 caption = "FSDT 2017; Fig 5.2",
                       x = "Savings/Investment Income",
                       y = "Percent")
gender_income_bargraph

```
<br>
Distribution of Respondents' Education Level
<br>
```{r}
#Generate the required data
educ_data = data %>%
  select(ID, Q1, Q4) %>% 
  rename(Age = Q1,Education_Level = Q4) %>%
  mutate(educ_level = if_else(Education_Level == 1, "No formal education",
                      if_else(Education_Level == 2, "Some Primary",
                      if_else(Education_Level == 3, "Primary completed",
                      if_else(Education_Level == 4, "Post primary technical training",
                      if_else(Education_Level == 5, "Some secondary",
                      if_else(Education_Level == 6, "University or other higher education","Don’t know"))))))) %>%
  group_by(educ_level) %>%
  summarise(educ_count = n())
educ_data

#Generate the bargraph(with the bars ordered)
educ_bargraph = ggplot(data = educ_data, 
                       aes(x = educ_level, 
                           y = educ_count))+
                geom_bar(stat = "identity", fill = "purple")+
                geom_text(aes(label = educ_count, hjust = -0.3, 
                              vjust = 0.5))+
                bargraph_theme+
                coord_flip()+
                scale_y_continuous(limits = c(0,3500), breaks = seq(0,3500,
                                                                    by=500))+
                labs(title = "Distribution of Respondents' Education Level",
                     caption = "FSDT 2017;Fig 5.3",
                     x = "Education Level",
                     y = "Frequency")
educ_bargraph

```
